Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-11-09 analysis/paper/paper.Rmd
Overhauled text, code changes and more

Attempting to define what constitutes a *quartzite* is not so stratightforward. Prieto et al. [-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. The defintion of quartzite has been a topic of discussion in geology as well [@howard05], and it hence varies somewhat between authors. Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard also emphasises the need for field-based methodologies. There are also situations where destructive techniques simply are not applicable. Artefact collections (museum, private, etc.) represent a valuable resource, as well as one of the easier ways of gaining access to large datasets with some geographical distribution. While destructive methods may prohibit access to such material there are still non-destructive methods that can generate valuable information. Aside from being non-destructive spectroscopic instrumentation feature fast measurements, making it an attractive alternative for field-based screening purposes. While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11; @tykot03], due to its capability of providing compositional data, there are other instrumentations that may provide a deeper insight into the structural properties. The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto19]. The same can be said for Raman spectroscopy, which has been applied extensively in the Cultural Heritage sector and in the analysis of artwork and manuscripts [@vandenabeele12].

In the current study a dataset of quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman, XRF). The goal is to evaluate their capabilities for fingerprinting quartz and quartzite in the North Swedish region. This will be achieved via Principal Component Analysis (PCA), where the observed groupings of the material in the model will be related to the spectral features for each instrumentation. 
 \linebreak

@@

The current study is based on spectroscopic surface measurements on quartz and quartzite points recovered from sites within the County of Västerbotten, Sweden. A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz/quartzite material stored at the collections at Västerbotten (www.vbm.se) and Skellefteå (www.skellefteamuseum.se) museums were analysed using NIRS, raman and XRF spectroscopy (Appendix XX). The material features a mainly black (N1-N3), grey or white (N9) hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent. The colour of the material has been classified both according to the munsell system, as well as sorted into general categories of hue (*dark* ~ N1-N4, *light* ~ N5-N8, *white* ~ N9, *Colourless* ~ lack of colour, opaque to translucent) for visualization purposes. A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region, previous research suggest they were in use during the 4 000 – 2 000 BP period [@forsberg85; @forsberg89; @mjaerum12]. Similarly, early studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite compared to previous periods [@baudou78].

@@

Västerbotten museum have come far in their digitization effort, making large portions of their collections searchable online. Based on a query of all points labelled as *quartz*, *brecciated quartz* or *quartzite* a list was generated from their database. Sampling of this pool of artefacts was not random, and based on geographical distribution, in an effort to include as many artefacts as possible from areas with few finds available, such as along the Vindel river. None of these were from the northernmost Skellefte river, however, and thus a selection of 5 artefacts each from three well-known sites stored at the collections of Skellefte museum were included. This resulted in a total dataset of `r length(tot_uniq)` artefacts, and after documentation and looking up missing information, `r length(sample_uniq)` of these were determined to be either points or preforms made from quartz/quartzite material.

@@

```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="White) Sites with measured samples in the County of Västerbotten; Black) Parish centroid for measured samples lacking site coordinates. The distribution of sampled sites is likely affected by the excavation and survey history of the region. Made with Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
knitr::include_graphics('../figures/Distribution.jpeg', error = FALSE)
```

@@

While parts of the collections at Västerbotten museum have been petrographically analysed by a geologist [@biornstad68; @ahman67], at this point it is difficult to say with any certainty how many (if any) of the artefacts in the current dataset were among these. It is fair to assume that the vast majority have been classified visually based on megascopic features according to a “standard” archaeological nomenclature. In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz and brecciated quartzite^[Original Swedish terms are *brecciekvarts*/*brecciekvartsit*]. Those that had no prior material classification registered were sorted into either the quartz or quartzite category. The reason for including *quartz* in the study is in order to test whether it is possible to separate quartz from quartzite using these methods, as well as a way to account for potential mistakes in the original classification.

Another reason for keeping a *quartz* category is the group referred to as *brecciated quartz*. In a petrographic study of the artefacts recovered along the Ume river valley Åhman (1967:8) briefly touches upon the definition of *brecciated quartz*, also referring to it as a *mylonite quartz*. Åhman describes the material as having formed from a quartz vein which has fractured due to movements in the mountain massives and recrystallized via pressure. While the translucent material may give the impression of an unusually pure quartzite, Åhman states that the thin-section analysis reveals the fragmented configuration of quartz. *Brecciated quartz* is a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g. @holm91].

@@

\linebreak
 \linebreak
All measurements were performed at the Environmental Archaeology Lab at Umeå University. The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens. Each artefact was measured twice on different sides, with additional measurements in cases where sampling was difficult. The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may be of the same material, however, for the smaller sites there still remains some uncertainties. For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.

Pre-processing of the spectra is different for each instrumentation and based on relevant literature. After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset [@geladi_linderholm20]. PCA modelling was performed using the base R *prcomp* function [@Rteam17; @rcoreteam21].

@@

NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe. The ASD features a detection range of 350 nm – 2500 nm, with spectral sampling (resp. spectral resolution) of 1.4 nm (resp. 3 nm) in the visible and near infrared range and 1.1 nm (resp. 10 nm) in the short‐wave infrared range. Each object was measured against a white reference background and the collected measurements were averaged for each artefact. Two parts of the As there were no notable features in the visible range the analysis was restricted to the 1 000 nm – 2 500 nm region in order to lessen the impact of noise. The data was mean-centred before PCA modelling.

@@

## Raman
Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties [@nafie01]. However, where NIRS generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both Rayleigh scatter and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01].

The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 cm^-1^ – 2 500 cm^-1^ ^[Raman shifts are typically given in wavenumber (cm^-1^), instead of spectral wavelength (nm)] with a resolution of 9.5 cm^-1^. Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.

@@

## X-Ray Fluourescence
XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand. The reference calibration used for element quantification was set to *mining* mode. Sampling with this setting takes 2 minutes in order for the instrument to cycle through all of the filters (main, low, high, light). Due to time constraints and the number of objects in the current dataset it was only possible to collect one measurement on each object, apart from some exceptions. The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling. As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL, or missing values, during PCA. Since missing values are not permitted during PCA these had to imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data. Two methods for imputation can be implemented, iterative or regularized, and since the former suffers when there is a large number of missing values the latter was used.

@@

## NIR - visual range

The main explanatory power for the visual region can be found in the first component with an explained variance of 98.4 %. In the PCA (fig. \@ref(fig:vis-nir-pca)) there is a clear trend along PC1 with darker materials being grouped together to the right of the plot and light/colourless/white material to the left. Examining the average spectra of these groups, as expected, the darker material feature higher absorbance of light than the other groups. No other distinctive features can be seen in the visual band, that could potentially be used to distinguish the groups. It is, however, apparent that while the more extreme hues of *dark* and *white* material can be isolated more clearly based on their spectra, there also exists a wider greyscale of material that does not conform as easily to these broader groups. There are also a number of artefacts with a dark hue that gravitate closer to the lighter material in the PCA. While the small size of these artefacts could be an explanation for the placement of these within the model, there are plenty of similarly sized artefacts that do not follow the same trend.

There are also instances where the measurements on different sides of an artefact end up in different groups in the PCA. In examination of the non-averaged data artefact 403 and 502 both have a sample taken on one side in the *dark* group and the other in the *greyish* middle. This is most likely due to the heterogeneous character of the material and natural variations on the surface possibly related to weathering. The *greyish* middle group in the PCA is otherwise largely made up of opaque material with a beige/grey tint.

@@

```{r vis-nir-pca, fig.width=20, fig.height=12, fig.cap="PCA score plot of PC1 and PC2 for the 400 nm – 750 nm band of the spectrum. Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant. The strong trend along PC1 also makes it difficult to separate out the muddled grayscale in the middle."}

@@

#prepare labels for PCs
vis.pc1var <- round(summary(visnir.pca)$importance[2,1]*100, digits=1)
vis.pc2var <- round(summary(visnir.pca)$importance[2,2]*100, digits=1)
vis.pc3var <- round(summary(visnir.pca)$importance[2,3]*100, digits=1)
vis.pc1lab <- paste0("PC1 (",as.character(vis.pc1var),"%)")
vis.pc2lab <- paste0("PC2 (",as.character(vis.pc2var),"%)")
vis.pc3lab <- paste0("PC3 (",as.character(vis.pc3var),"%)")

@@

```{r vis-nir-spectra, fig.width=20, fig.height=15, fig.cap="Average NIR spectra by hue. Darker samples feature higher overall absorbance with white material absorbing the least amount of light."}

#Filter NIR data to focus on the material with a dark hue 
#and select the NIR range 1 000 - 2 500 nm
vis.dark <- Points.nir %>% 
  filter(hue == "Dark") %>% 
  select(c(1:28),c(79:429))

spec.vd <- 
      as_spectra(vis.dark, name_idx = 1, meta_idxs = 2:28)

vis.light <- Points.nir %>% 
  filter(hue == "Light") %>% 
  select(c(1:28),c(79:429))

spec.vl <- 
  as_spectra(vis.light, name_idx = 1, meta_idxs = 2:28)

vis.colourless <- Points.nir %>% 
  filter(hue == "Colourless") %>% 
  select(c(1:28),c(79:429))

spec.vt <- 
  as_spectra(vis.colourless, name_idx = 1, meta_idxs = 2:28)

vis.white <- Points.nir %>% 
  filter(hue == "White") %>% 
  select(c(1:28),c(79:429))

spec.vw <- 
  as_spectra(vis.white, name_idx = 1, meta_idxs = 2:28)

par(mfrow=c(4,1))

plot(mean(spec.vd), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark"); vps1 <- do.call(vpStack, baseViewports())

plot(mean(spec.vl), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Light")

plot(mean(spec.vt), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Colourless")

plot(mean(spec.vw), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White"); vps4 <- do.call(vpStack, baseViewports())
```

@@

## NIR - near infrared range
In the case of NIRS similar features can be observed in the average spectra for all four colour groups, with darker materials exhibiting higher absorbance and the light group the most distinct peaks. The presence of molecular water can be seen in the peaks at 1 400 nm and 1 900 nm  [@hunt77; @sciuto_etal19] (fig. \@ref(fig:nir-spectra)). Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter falls within one of the stronger CO^3^ bands [@clark_etal90]. Inspection of individual spectra for the *light* group, however, also reveals a 5th peak for some samples at 2 450 nm. Comparisons to spectra collected on muscovite ([@clark_etal90] p.12 675) shows similar peaks located at 2 200 nm, 2 350 nm, and 2 450 nm. It is therefore possible that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other samples Al-OH.

The PCA shows a strong trend along PC 1 that seems to correlate with the colour of the material(fig. \@ref(fig:nir-pca)). Darker quartzite cluster to the right with white and colourless samples grouping to the opposite side. While the white samples seem to gravitate to the bottom left of the model the samples that are somewhat translucent, such as the *brecciated quartz*, seem to cluster towards the top-left. A small set of 3 black samples at the top of the model separate themselves clearly from the rest of the data points. These three samples come from different sites and exhibit no peaks in their spectra, making them an interesting set of outliers.

@@

```{r nir-pca, fig.width=20, fig.height=8, fig.cap= "PCA score plot of 1st and 2nd components for the 1 000 nm – 2 500 nm region of the spectrum. Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left quadrant. A trend of translucent material and opaque material with a beige tint grouping in the top left can also be seen. Three black samples at the top, from different sites, all feature no distinguishing peaks in their spectra thus separating them from the other samples in the dataset."}

@@

#prepare labels for PCs
nir.pc1var <- round(summary(nir.pca)$importance[2,1]*100, digits=1)
nir.pc2var <- round(summary(nir.pca)$importance[2,2]*100, digits=1)
nir.pc1lab <- paste0("PC1 (",as.character(nir.pc1var),"%)")
nir.pc2lab <- paste0("PC2 (",as.character(nir.pc2var),"%)")

@@

## Raman
There are two distinctive features present in the Raman data (fig. \@ref(fig:raman-spectra)). The first are three peaks in the 100 cm^-1^ - 500 cm^-1^ band that are typical for SiO~2~ and quartz-based materials, with the strongest peak occuring at 460 cm^-1^ [@gillet_etal90]. These peaks are present for all samples in the dataset. The second feature shows up as two peaks occurring at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite/graphene [@cesare_maineri99]. The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@vollebregt_etal12; @lee_etal21]. Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm^-1^ region. It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.

>>>>

Prieto et al. [-@prieto_etal19 p.15; -@prieto_etal20 p.32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature, and different researchers have thus emphasized different features as being characteristic of the material [@howard05]. Visual characterization on a megascopic level is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05 p.708] also emphasises the need for field-based methodologies. In the case of archaeological research it is worth considering situations where destructive techniques simply are not applicable. Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution. While destructive methods may prohibit access to such material there are still non-destructive alternatives that can generate valuable information. Aside from being non-destructive spectroscopic instrumentation feature fast measurements, making it an attractive option for field-based screening purposes. While X-Ray Fluorescence (XRF) is the go-to spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing compositional data, there are other instrumentations that may provide a deeper insight into the structural properties. The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF [@hunt77], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@linderholm_geladi14a; @linderholm_geladi14b; @sciuto19]. The same can be said for Raman spectroscopy, which has been applied extensively in the Cultural Heritage sector and in the analysis of artwork and manuscripts [@vandenabeele12].

In the current study a dataset of quartz and quartzite artefacts from museum collections in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman, XRF). The aim is to evaluate their viability for fingerprinting quartz and quartzite material in the North Swedish region, where there currently is a lack of reference material as well as a complex geological setting. This will be achieved via Principal Component Analysis (PCA), where the observed groupings of the material in the model will be related to their spectral features. 
 \linebreak

@@

The current study is based on spectroscopic surface measurements collected on quartz and quartzite points recovered from sites within the County of Västerbotten, Sweden. A dataset of `r length(sample_uniq)` points, point fragments and preforms made from quartz and quartzite material stored at the collections at Västerbotten (www.vbm.se) and Skellefteå (www.skellefteamuseum.se) museums were analysed using NIRS, raman and XRF spectroscopy. The material features a mainly black, grey or white hue, as well as a subset of semi-translucent material with a beige tint to almost fully translucent. The colour of the material has been classified both according to the Munsell system, as well as sorted into general categories of hue for visualization purposes: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), *colourless* (lack of colour, opaque to translucent). A number of these points have, according to prior classifications, been worked using bifacial technology. While there currently is no local chronology available for this artefact type in the North Swedish region, previous research suggest they were in use during the 4 000 – 2 000 BP period [@forsberg85; @forsberg89; @mjaerum12]. Similarly, early studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite compared to previous periods [@baudou78].

@@

Västerbotten museum have come far in their digitization effort, making large portions of their collections searchable online. Based on a query of all points and preforms labelled as *quartz*, *brecciated quartz* or *quartzite* a list was generated from their database. Sampling of this list of artefacts was not random. Sites located in areas with few finds were included first in order to increase the geographical distribution, with sampling of areas with higher site density making up the remaining bulk of the dataset. None of these were from the northernmost Skellefte river, however, and thus a selection of 5 artefacts each from three well-known sites stored at the collections of Skellefte museum were included. This resulted in a total dataset of `r length(tot_uniq)` artefacts, and after documentation and looking up missing information, `r length(sample_uniq)` of these were determined to be either points or preforms made from quartz or quartzite material.

@@

```{r distribution-map, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Geographical distribution of sampled sites within Västerbotten County, Sweden. The distribution of sampled sites is likely affected by the excavation and survey history of the region. White: Sampled sites; Black: Parish centroid for measured samples lacking site coordinates. Made with Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."}
knitr::include_graphics('../figures/Distribution.jpeg', error = FALSE)
```

@@

While parts of the collections at Västerbotten museum have been petrographically analysed by a geologist [@biornstad68; @ahman67], at this point it is difficult to say with any certainty how many (if any) of the artefacts in the current dataset were among these. It is fair to assume that the vast majority have been classified visually based on megascopic features according to a “standard” archaeological nomenclature. In the dataset the four main types of registered materials are: quartz, quartzite, brecciated quartz and brecciated quartzite^[Original Swedish terms are *brecciekvarts*/*brecciekvartsit*]. Those that had no prior material classification registered were sorted into either the quartz or quartzite category. The reason for including quartz in the study is in order to test whether it is possible to separate quartz from quartzite using these methods, as well as a way to account for potential mistakes in the original classification.

Another reason for keeping a quartz category is the group referred to as *brecciated quartz*. In a petrographic study of the artefacts recovered along the Ume river valley Åhman [-@ahman67 p.8] briefly touches upon the definition of *brecciated quartz*, also referring to it as a *mylonite quartz*. Åhman describes the material as having formed from a quartz vein which has fractured due to movements in the mountain massives and recrystallized via pressure. While the translucent material may give the impression of an unusually pure quartzite, Åhman states that the thin-section analysis reveals the fragmented configuration of quartz. *Brecciated quartz* is a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g. @holm91 p.24].

@@

All measurements were performed at the Environmental Archaeology Lab at Umeå University^[www.umu.se/forskning/infrastruktur/mal/]. The analysis was conducted in a dark room in order to minimize light pollution, with the main light sources in the room being the computer screens. Each artefact was measured twice, on different sides, with additional measurements in cases where sampling was difficult. The only exception to this was the XRF, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the statistical reliability of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may be of the same material, however, for the smaller sites there still remains some uncertainties. For materials that featured high heterogeneity the sample point was focused, as much as possible, on the surrounding matrix.

Pre-processing of the spectra is different for each instrumentation and based on relevant literature. After pre-processing the data was modelled using PCA, which is an unsupervised exploratory technique for reducing the dimensionality of the dataset [@geladi_linderholm20]. PCA modelling was performed using the base R *prcomp()* function [@Rteam17; @rcoreteam21].

@@

NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe. The ASD features a detection range of 350 nm – 2 500 nm, with spectral sampling (resp. spectral resolution) of 1.4 nm (resp. 3 nm) in the visible and near infrared range and 1.1 nm (resp. 10 nm) in the short‐wave infrared range. Each object was measured against a white reference background and the collected measurements were averaged for each artefact. The visible spectrum (~ 400 nm - 750 nm) and the NIR range (~ 1 000 nm – 2 500 nm) were analysed separately, with the former focused on colour analysis and the latter the chemical structure of the material. The data was mean-centred before PCA modelling.

@@

## Raman
Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties [@nafie01]. However, where NIRS measures light absorbance and generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both Rayleigh scatter and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01].

The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 cm^-1^ – 2 500 cm^-1^ ^[Raman shifts are typically given in wavenumber (cm^-1^), instead of spectral wavelength (nm)] with a resolution of 9.5 cm^-1^. Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå^[www.slu.se/en/departments/forest-biomaterials-technology/].

@@

## X-Ray Fluourescence
XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method for petrographic analysis in archaeology [@hall60; @shackley11a]. Exposing a material to X-ray radiation causes electrons to dislodge and move between the electron shells of the atoms, which in turn emits energy (i.e. *fluourescence*). As the energy differences between electron shells for the various elements is known it is possible to both identify and quantify the abundance of an element [@shackley11b p.16], which is useful for provenience studies where the chemical character of an unknown and a reference sample can tested statistically [@tykot03].

XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand. The reference calibration used for element quantification was set to *mining* mode. Sampling with this setting takes 2 minutes in order for the instrument to cycle through all of the filters (main, low, high, light). Due to time constraints and the number of objects in the current dataset it was only possible to collect one measurement on each object, apart from some exceptions.

The XRF data was log-transformed, mean-centered and scaled to unit variance before PCA modelling. As the resulting dataset also features values below the Limit of Detection (LOD), these were treated as NULL (or "missing values"), during PCA. Since missing values are not permitted during PCA these had to imputed before the analysis, which was performed using the *missMDA* R package [@josse_husson16] developed for datasets with incomplete data. Two methods for imputation can be implemented, iterative or regularized, and since the former suffers when there is a large number of missing values the latter was used.

@@

## NIR - visual range
Examining the average spectra of the different colour groups (dark, light, white and colourless), as expected, the darker material feature higher absorbance of light than the other groups. No other distinctive features can be seen in the visual band, that could potentially be used to distinguish them. It is, however, apparent that while the more extreme hues of dark and white material can be isolated more clearly based on their spectra, there also exists a wider greyscale of material that does not conform as easily to these broader groups.

The cumulative explained variance for the first three components in the PCA is ~ 99.9%, with the accompanying loadings in fig. \@ref(fig:vis-nir-loadings). Most of the contribution to the first Principal Component (PC) can be found within the red part of the spectrum (~ 600 nm - 750 nm). This can be seen in the score plot where darker and lighter materials are separated along PC 1, as darker samples would feature a higher absorbance within the red part of the spectrum (fig. \@ref(fig:vis-nir-pca)). In contrast, most of the contribution to PC 2 occurs within the blue-violet part of the spectrum, which does not seem to alter the structure of the dataset in any significant way. PC 3 feature high contributions within both the blue-violet and red parts of the spectrum, which also seems to draw the light and dark samples that feature a blue-gray tint towards the top of the score plot.

While the overall structure of the model seems to group colourless and white materials towards the bottom left, with dark material to the right, there are a number of samples that break this pattern. This suggests that the perceived colour of a material does not necessarily match what is detected within the visible spectrum of light. It is also difficult to separate out the greyscale of colours that occur inbetween the "black and white" groups, where both light gray and semi-translucent samples end up. All in all, the results echo the complexity and unreliability that exists within the topic of petrographic colour.

@@

```{r vis-nir-pca, fig.width=20, fig.height=12, fig.cap="PCA score plots for the 400 nm – 750 nm band of the NIR data. Top: PC 1 and PC 2, bottom: PC 1 and PC 3 PC 1 seems to group dark-light material on opposite sides of the plots while PC 2 does little to change the strucure. PC 3 seems to position material with a blue-gray tint towards the top of the score plot."}

@@

#prepare labels for PCs
vis.pc1var <- round(summary(visnir.pca)$importance[2,1]*100, digits=2)
vis.pc2var <- round(summary(visnir.pca)$importance[2,2]*100, digits=2)
vis.pc3var <- round(summary(visnir.pca)$importance[2,3]*100, digits=2)
vis.pc1lab <- paste0("PC1 (",as.character(vis.pc1var),"%)")
vis.pc2lab <- paste0("PC2 (",as.character(vis.pc2var),"%)")
vis.pc3lab <- paste0("PC3 (",as.character(vis.pc3var),"%)")

@@

```{r vis-nir-loadings, fig.width=20, fig.height=15, fig.cap="Loading lines for PC 1 - 3. The two first components mainly describes features with a linear increase in absorbance across the visual spectrum, but the 3rd component demonstrates more variation, with lower absorbance within the blue-yellow region. "}

#Filter NIR data to focus on the material with a dark hue 
#and select the NIR range 1 000 - 2 500 nm
#vis.dark <- Points.nir %>% 
#  filter(hue == "Dark") %>% 
#  select(c(1:28),c(79:429))

#spec.vd <- 
#      as_spectra(vis.dark, name_idx = 1, meta_idxs = 2:28)

#vis.light <- Points.nir %>% 
#  filter(hue == "Light") %>% 
#  select(c(1:28),c(79:429))

#spec.vl <- 
#  as_spectra(vis.light, name_idx = 1, meta_idxs = 2:28)

#vis.colourless <- Points.nir %>% 
#  filter(hue == "Colourless") %>% 
#  select(c(1:28),c(79:429))

#spec.vt <- 
#  as_spectra(vis.colourless, name_idx = 1, meta_idxs = 2:28)

#vis.white <- Points.nir %>% 
#  filter(hue == "White") %>% 
#  select(c(1:28),c(79:429))

#spec.vw <- 
#  as_spectra(vis.white, name_idx = 1, meta_idxs = 2:28)

#VISNIR LOADING LINES
#For some reason invisible() has to be used to prevent console output showing up in knitted document
par(mfrow=c(3,1))
invisible(sapply(1:3, function(comp){
  plot(400:750, visnir.pca$rotation[,comp], type='l', lwd=2,
      main=paste("PC", comp), xlab="Wavelength (nm)", ylab="", cex.lab = 1.2, xaxp=c(400,750,5))}))

#par(mfrow=c(4,1))

#plot(mean(spec.vd), lwd = 1.2, col = "black", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Dark"); vps1 <- do.call(vpStack, baseViewports())

#plot(mean(spec.vl), lwd = 1.2, col = "red", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Light")

#plot(mean(spec.vt), lwd = 1.2, col = "purple", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - Colourless")

#plot(mean(spec.vw), lwd = 1.2, col = "blue", ylab = "Absorbance", cex.lab = 1.5, cex.main = 1.5, xlab = "Wavelength (nm)", main = "Average - White"); vps4 <- do.call(vpStack, baseViewports())
```

@@

## NIR - near infrared range
In the case of the 1 000 nm - 2 500 nm band of the NIR data, similar features can be observed in the average spectra for all four colour groups, with darker materials exhibiting higher absorbance and the light group the most distinct peaks. The presence of molecular water can be inferred based on the peaks at 1 400 nm and 1 900 nm  [@hunt77; @sciuto_etal19] (fig. \@ref(fig:nir-spectra)). Peaks also occur at 2 200 nm and 2 350 nm, where the former has been related to combination vibrations of Al-OH and the latter falling within one of the stronger CO^3^ bands [@clark_etal90]. Inspection of individual spectra for the light group, however, also reveals a 5th peak for some samples at 2 450 nm. Comparisons to spectra collected on muscovite [@clark_etal90 p.12 675] shows similar peaks located at 2 200 nm, 2 350 nm, and 2 450 nm. It is therefore possible that the 2 350 nm band in some cases corresponds to CO^3^ vibrations and in other samples Al-OH, possibly inidicating the presence of muscovite.

The PCA shows a strong trend along PC 1 that seems to correlate with the colour of the material(fig. \@ref(fig:nir-pca)). Darker quartzite cluster to the right with white and colourless samples grouping to the opposite side. While the white samples seem to gravitate to the bottom left of the model the samples that are somewhat translucent, such as the brecciated quartz, seem to cluster towards the top-left. At the top of the score plot 3 black samples separate themselves clearly from the rest of the data points. These three samples come from different sites and exhibit no peaks in their spectra, making them an interesting set of outliers.

@@

```{r nir-pca, fig.width=20, fig.height=8, fig.cap= "PCA score plot of 1st and 2nd components for the 1 000 nm – 2 500 nm region of the spectrum. Quartz-based material with a darker surface group to the right in the model, while those with white surface group in the bottom left. A trend of translucent material and opaque material with a beige tint grouping in the top left can also be seen. Three black samples at the top, from different sites, all feature no distinguishing peaks in their spectra thus separating them from the other samples in the dataset."}

@@

#prepare labels for PCs
nir.pc1var <- round(summary(nir.pca)$importance[2,1]*100, digits=2)
nir.pc2var <- round(summary(nir.pca)$importance[2,2]*100, digits=2)
nir.pc1lab <- paste0("PC1 (",as.character(nir.pc1var),"%)")
nir.pc2lab <- paste0("PC2 (",as.character(nir.pc2var),"%)")

@@

## Raman
There are two distinctive features present in the Raman data (fig. \@ref(fig:raman-spectra)). The first are three peaks in the 100 cm^-1^ - 500 cm^-1^ band that are typical for SiO~2~ and quartz-based materials, with the strongest peak occuring at 460 cm^-1^ [@gillet_etal90]. These peaks are present for all samples in the dataset. The second feature shows up as two peaks occurring at around 1 287 cm^-1^ and 1 598 cm^-1^, and dominate the spectra in terms of intensity for the darker samples. These peak positions fall within what is referred to as the *D* and *G* band respectively, and have been observed in spectroscopic studies of graphite [@cesare_maineri99]. The D band occurs as a result of disordered carbon (C) structures whilst the G band is indicative of sp^2^ carbon networks and caused by stretching in the carbon bonds, which is common in graphite-based materials [@cohen-ofri_etal06; @vollebregt_etal12; @lee_etal21]. Whereas the 2^nd^ peak falls within the typical position of the G band, the position of the D band is commonly identified as being in the 1 330 - 1 350 cm^-1^ region. It has, however, been documented that the dispersive behavior of the D band can cause a frequency change as a result of the energy of the incident laser [@pimenta_etal07 p.1280]. This could thus explain the discrepancy between the observed peak in the dataset and the literature.
